---
title: "Grid Fundamentals & Concepts"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Grid Fundamentals & Concepts}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(vaster)
```

## What is vaster?
 
The **vaster** package provides pure raster grid logic without requiring actual raster data. It separates the abstract mathematical operations of raster grids—dimension, extent, cell indexing, coordinate transformations—from data formats and file I/O.
 
This is useful when you need to:

- Pre-compute grid operations before reading or writing data
- Work with virtual grids and mosaics
- Perform format-agnostic grid calculations
- Build higher-level geospatial tools

## The Virtual Grid Model

A raster grid in vaster is defined by just two components:

1. **Dimension**: the number of columns and rows `c(ncol, nrow)`
2. **Extent**: the bounding box `c(xmin, xmax, ymin, ymax)`

```{r}
## a 10x5 grid covering [0, 100] x [0, 50]
dimension <- c(10, 5)
extent <- c(0, 100, 0, 50)
```

From these two pieces of information, all grid properties can be derived:

```{r}
n_cell(dimension)
x_res(dimension, extent)
y_res(dimension, extent)
```

There is no data, no file format, no CRS—just the logical grid structure. This matches how GDAL and other tools internally represent grids.

## Cell Numbering Convention

Cells are numbered from 1, starting at the **top-left** corner and proceeding **row-wise** (left-to-right, top-to-bottom). This matches the convention used by the `raster` and `terra` packages.
 
```{r}
## cell numbers for a 4x3 grid
dimension <- c(4, 3)
extent <- c(0, 4, 0, 3)

## total cells
n_cell(dimension)

## cell 1 is top-left
xy_from_cell(dimension, extent, 1)

## cell 4 is top-right  
xy_from_cell(dimension, extent, 4)

## cell 12 is bottom-right
xy_from_cell(dimension, extent, 12)
```

The cell centre coordinates reflect this ordering:

```{r}
## all cell centres
cells <- seq_len(n_cell(dimension))
xy_from_cell(dimension, extent, cells)
```

## Resolution and Origin

**Resolution** is the cell size in x and y directions:

```{r}
dimension <- c(10, 5)
extent <- c(0, 100, 0, 50)

x_res(dimension, extent)  
y_res(dimension, extent)  
```

The **origin** is the point where cell boundaries align. By default, vaster uses the bottom-left corner of the extent:

```{r}
origin(dimension, extent)
```

Understanding the origin is important for aligning multiple grids—covered in detail in `vignette("extent-alignment")`.

## Extent Properties

Extract individual extent components, dimension is obviously irrelavent for extent components but
is included for consistency with other functions. 
 
```{r}
extent <- c(-180, 180, -90, 90)

x_min(c(1, 1), extent)
x_max(c(1, 1), extent)
y_min(c(1, 1), extent)
y_max(c(1, 1), extent)

xlim(c(1, 1), extent)
ylim(c(1, 1), extent)
```

## Coordinate Arrays

Generate coordinate vectors for all columns or rows:

```{r}
dimension <- c(4, 3)
extent <- c(0, 4, 0, 3)

## x coordinates of cell centres (one per column)
x_centre(dimension, extent)

## y coordinates of cell centres (one per row)
y_centre(dimension, extent)

## x coordinates of cell corners/edges (ncol + 1 values)
x_corner(dimension, extent)

## y coordinates of cell corners/edges (nrow + 1 values)
y_corner(dimension, extent)
```

For a full matrix of all cell centre coordinates:

```{r}
xy(dimension, extent)
```

## Comparison with raster/terra

The vaster approach differs from `raster` and `terra` in that there is no object to create or manage. Grid operations are pure functions:

```{r, eval = FALSE}
## terra approach
library(terra)
r <- rast(ncol = 10, nrow = 5, xmin = 0, xmax = 100, ymin = 0, ymax = 50)
res(r)
ncell(r)
xyFromCell(r, 1)

## vaster approach
dimension <- c(10, 5)
extent <- c(0, 100, 0, 50)
c(x_res(dimension, extent), y_res(dimension, extent))
n_cell(dimension)
xy_from_cell(dimension, extent, 1)
```

This functional approach is useful when you need to perform many grid calculations without the overhead of object creation, or when working in contexts where raster packages aren't available.

## Next Steps

- `vignette("cell-operations")`: Converting between cells, coordinates, and row/col indices
- `vignette("extent-alignment")`: Aligning and cropping extents to grids
- `vignette("gdal-interop")`: Working with GDAL geotransforms and RasterIO
